Chess EDA

### 2023-04-17 owen/Write Ups/01_eda.qmd
fixing a bug in the model where the previous n games included the current result we are predicting. This explains the problem with all the effects looking the same. No gives more reasonable results, permuting the data gives clear indications for no effects then also

---
title: "Chess EDA"
author: Owen G. Ward
format: 
  html:
    toc: true
    toc-location: left
bibliography: refs.bib
---

@@

lichess_data <- readRDS("../rdata/lichess_pilot.RData")
```

@@

Similar to @ding2021nhl, we wish to fit a multilevel logistic 
regression to see if we can identify individual 
effects for chess players in this data.
We will incorporate fixed effects for the player effects of the focal
player in these games, along

We will identify focal players as those who were used to query the API,
i.e the `Username` for each game. We only consider Rated 60 second bullet games
here, and use the top 18 players, who all play at least 100 games
of 1 minute rated bullet chess in this dataset.

@@

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, top_players)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = focal_win_prop - mean(focal_win_prop)) %>% 
  filter(focal_result != 0.5)

@@

$$
\alpha_j \sim \mathcal{N}(\mu_1, \tau_1)
$$
$$
\mu_1 \sim \mathcal{N}(0, 5)
$$
$$
\tau_1 \sim \mathcal{Cauchy}(0, 5)
$$

@@

$$
\beta_j \sim \mathcal{N}(\mu_2, \tau_2)
$$
$$
\mu_2 \sim \mathcal{N}(0, 5)
$$
$$
\tau_2 \sim \mathcal{Cauchy}(0, 5)
$$

@@

stan_file <- "model3.stan"

@@

fit3_ave$save_object("init_model_n10.RDS")
```

@@

fit3_ave <- readRDS("init_model_n10.RDS")

@@

Similarly, $\gamma_2$ tells us about how the ELO of the focal player 
being larger than their opponent influences their win probability.
The estimate here indicates that a large ELO difference makes a player very
likely to win, as would be expected. For example, a player who
has a 50% chance of winning against an opponent with the same ELO
will have about a 56% of winning against an opponent with an ELO score
100 points lower.

@@

fit3_ave$save_object("init_model_n5.RDS")

@@

```{r}
fit3_ave <- readRDS("init_model_n5.RDS")

@@

fit3_ave$save_object("init_model_n20.RDS")

@@

fit3_ave <- readRDS("init_model_n20.RDS")

@@

fit3_ave$save_object("init_model_n50.RDS")
```

```{r}
#| message: false
fit3_ave <- readRDS("init_model_n50.RDS")

@@

Using a longer history seems to give a model which fits quicker and
is more stable. But I'm not sure why that should be the case?

@@

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, blitz_top)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = focal_win_prop - mean(focal_win_prop)) %>% 
  filter(focal_result != 0.5)

@@

stan_file <- "model3.stan"

@@

fit3_ave$save_object("init_model_blitz_n5.RDS")

@@

```{r}
fit3_ave <- readRDS("init_model_blitz_n5.RDS")

@@

stan_file <- "model3.stan"

@@

fit3_ave$save_object("init_model_blitz_n20.RDS")

@@

```{r}
fit3_ave <- readRDS("init_model_blitz_n20.RDS")

@@

@@

- There are some potential concerns with this Stan model and that it could
be made more efficient and better parameterized. These seem to be a
larger issue when the number of previous games used for the win ratio
is smaller, which would make sense as these would depart more from the mean.

@@

@@

```{r}
#| echo: true
prev_data <- bullet_60 %>% 
  group_by(Username) %>% 
  arrange(UTCDate, UTCTime) %>% 
  ## get focal result here also
  mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
  mutate(focal_result = case_when(
    (focal_white == 1 & Result == "1-0") ~ 1,
    (focal_white == 0 & Result == "0-1") ~ 1,
    (Result == "1/2-1/2") ~ 0.5,
    .default = 0
    )) %>% 
  mutate(prev_result = lag(focal_result)) %>% 
  select(White, Black, WhiteElo, BlackElo, focal_white:prev_result)

>>>>

---
title: "Chess EDA"
author: Owen G. Ward
format: 
  # html:
  #   toc: true
  #   toc-location: left
  pdf
bibliography: refs.bib
---

@@

lichess_data <- readRDS("../../rdata/lichess_pilot.RData")
```

@@

Similar to @ding2021nhl, we wish to fit a multilevel logistic 
regression to see if we can identify individual 
effects for chess players in this data.
We will incorporate random effects for the player effects of the focal
player in these games, along with random player level winner effects. 
We take common fixed effects for the effect of colour and 
the ELO difference.

We will identify focal players as those who were used to query the API,
i.e the `Username` for each game. We only consider Rated 60 second bullet games
here initially, and use the top 18 players, who all play at least 100 games
of 1 minute rated bullet chess in this dataset.

@@

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, top_players)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = lag(focal_win_prop, default = 0) - mean(focal_win_prop)) %>% 
  filter(focal_result != 0.5)

@@

$$
\alpha_j \sim \mathcal{N}(\mu_2, \tau_2)
$$
$$
\mu_1 \sim \mathcal{N}(0, 5)
$$
$$
\tau_1 \sim \mathcal{Cauchy}(0, 5)
$$

@@

$$
\beta_j \sim \mathcal{N}(\mu_1, \tau_1)
$$
$$
\mu_2 \sim \mathcal{N}(0, 5)
$$
$$
\tau_2 \sim \mathcal{Cauchy}(0, 5)
$$

@@

stan_file <- "../model3.stan"

@@

fit3_ave$save_object("../model_fits/init_model_n10.RDS")
```

@@

fit3_ave <- readRDS("../model_fits/init_model_n10.RDS")

@@

Similarly, $\gamma_2$ tells us about how the ELO of the focal player 
being larger than their opponent influences their win probability.
The estimate here indicates that a large ELO difference makes a player very
likely to win, as would be expected. For example, a player who
has a 50% chance of winning against an opponent with the same ELO
will have about a 61% chance of winning against an opponent with an ELO score
100 points lower.

@@

fit3_ave$save_object("../model_fits/init_model_n5.RDS")

@@

```{r}
fit3_ave <- readRDS("../model_fits/init_model_n5.RDS")

@@

fit3_ave$save_object("../model_fits/init_model_n20.RDS")

@@

fit3_ave <- readRDS("../model_fits/init_model_n20.RDS")

@@

fit3_ave$save_object("../model_fits/init_model_n50.RDS")
```

```{r}
#| message: false
fit3_ave <- readRDS("../model_fits/init_model_n50.RDS")

@@

@@

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, blitz_top)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = lag(focal_win_prop, default = 0) - mean(focal_win_prop)) %>% 
  filter(focal_result != 0.5)

@@

# stan_file <- "../model3.stan"

@@

fit3_ave$save_object("../model_fits/init_model_blitz_n5.RDS")

@@

```{r}
fit3_ave <- readRDS("../model_fits/init_model_blitz_n5.RDS")

@@

# stan_file <- "model3.stan"
# 
# mod <- cmdstan_model(stan_file)

@@

fit3_ave$save_object("../model_fits/init_model_blitz_n20.RDS")

@@

```{r}
fit3_ave <- readRDS("../model_fits/init_model_blitz_n20.RDS")

@@

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1","gamma1", "gamma2")))
```

@@

@@

- How data dependent is this result? Should investigate the
performance across different players, the strength of the winner
effect as a function of the ELO of a player, etc.

@@

prev_data <- bullet_60 %>% 
  group_by(Username) %>% 
  arrange(UTCDate, UTCTime) %>% 
  ## get focal result here also
  mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
  mutate(focal_result = case_when(
    (focal_white == 1 & Result == "1-0") ~ 1,
    (focal_white == 0 & Result == "0-1") ~ 1,
    (Result == "1/2-1/2") ~ 0.5,
    .default = 0
    )) %>% 
  mutate(prev_result = lag(focal_result)) %>% 
  select(White, Black, WhiteElo, BlackElo, focal_white:prev_result)
